---
title: "Project title - exploratory data analysis"
author: 
  - Template by Bertalan Polner^[Donders Institute for Brain, Cognition and Behaviour, Radboud University, Nijmegen, The Netherlands]
  - You can add more authors^[and their affiliations]
date: "Last compiled on `r format(Sys.time(), '%d/%m/%Y, %X %Z')`"
output: 
  html_notebook:
    toc: true
    toc_float: true
    toc_depth: 5
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

This is a ***general*** template for tidying data and Exploratory Data.
Analysis (EDA), using mostly the tidyverse.

A very helpful introduction to EDA with tidyverse:
<https://r4ds.hadley.nz/eda>

# Setup

## Load libraries

It is good practice to load all libraries at the beginning of your
notebook.

```{r }
library(tidyverse) # very useful for importing, tidying, aggregating, and visualising data
```

## Global options for knitr

> To set global options that apply to every chunk in your file, call
> knitr::opts_chunk\$set in a code chunk. Knitr will treat each option
> that you pass to knitr::opts_chunk\$set as a global default that can
> be overwritten in individual chunk headers.
>
> Source: <https://rmarkdown.rstudio.com/lesson-3.html>

## Specify themes

You can use ggplot2::theme_set()
<https://ggplot2.tidyverse.org/reference/theme_get.html>

## Session info (to support reproducibility)

```{r}
sessionInfo()
```

# Import

-   A useful way to import multiple raw datafiles with identical format
    and merge them into a single dataframe
    -   Get all the relative file paths with list.files() - either have
        only the relevant files in a separate folder or specify a regex
        in the patter argument
    -   Import them with a proper file import function (usually
        read_csv).
    -   You may also specify import arguments and pass them to map_dfr
        or write your own import function for complex use cases
    -   Wrap the import within map_dfr: voila, you have all the raw data
        in a single dataframe!

```{r}
file_paths <- list.files("directory with raw data files here", full.names = TRUE)
dat        <- map_dfr(file_paths, read_csv) 
```

# Tidy

## Clean

Iterate between initial formal quality control and removing rows/columns
if necessary

### Initial formal quality control

Do some sanity checks:

-   right number of individuals (or sites, etc)?
-   right number of observations per individuals?
-   right number of columns?

```{r}
table(mtcars$mpg)
```
```{r}
summary(mtcars)
```


### Rows

Never change the raw data!

Are there unnecessary rows in your dataframe?

Try to formalize a rule that applies to them and use dplyr::filter to
get rid of them.

Avoid indexing by row number as that is not stable.

### Columns

Also, are there unnecessary columns in there?

There might be redundant columns automatically added by default during
data collection.

Refer to columns by name and not by position (again, that might become
unstable). Or use dplyr::select to select the columns you need and get rid of the rest. 

## Prepare

### Variable names

Are columns names intuitive and simple? If not, rename.

Avoid var2.rt or thisResponse.RT

Prefer reaction_time or rt (use abbreviations only when they are very
trivial!)

Are column names spelled consistently?

Sticking to a consistent style will save you mental capacity.

Hint: janitor package, tolower(), toupper(), etc.
<https://cran.r-project.org/web/packages/janitor/vignettes/janitor.html>

### Compute, recode, convert variables

You might realise that some variables can be represented in a better way
than done in the raw files. E.g. recode dummy variables to 0/1, or
convert categorical variables into factors. 

You may also want to calculate some variables that you will explore
later. It might make sense to do all those calculations at this point,
so they're available at all subsequent data exploration steps.

# Explore

> Your goal during EDA is to develop an understanding of your data. The
> easiest way to do this is to use questions as tools to guide your
> investigation. When you ask a question, the question focuses your
> attention on a specific part of your dataset and helps you decide
> which graphs, models, or transformations to make.
>
> EDA is fundamentally a creative process. And like most creative
> processes, the key to asking *quality* questions is to generate a
> large *quantity* of questions. It is difficult to ask revealing
> questions at the start of your analysis because you do not know what
> insights can be gleaned from your dataset. On the other hand, each new
> question that you ask will expose you to a new aspect of your data and
> increase your chance of making a discovery. You can quickly drill down
> into the most interesting parts of your data---and develop a set of
> thought-provoking questions---if you follow up each question with a
> new question based on what you find.
>
> There is no rule about which questions you should ask to guide your
> research. However, two types of questions will always be useful for
> making discoveries within your data. You can loosely word these
> questions as:
>
> 1.  What type of variation occurs within my variables?
>
> 2.  What type of covariation occurs between my variables?
>
> Source: <https://r4ds.hadley.nz/eda#questions>

It is good practice to look at aggregated and individual-level data as
well, to get a sense of both overall trends and individual variation.
Hints for the latter:

-   plotting: add a facet_wrap(\~participant) layer to plots
-   sumamry tables: add a group_by(participant) operation before summary
    stats

## Missing values

add naniar package

## Variation

Some general questions to get started:

-   are values within a reasonable range? Sanity check
-   is there enough variation?
-   what is the shape of the distribution?
-   what are the typical values?
-   are there unusual values? Outliers?
-   is there a pattern in which values are absent from the data?

## Covaration

Some general questions to get started:

-   do you see expected / trivial associations? Sanity check, again
-   do you see unexpected associations?
-   do you see clusters in the data?
-   multivariate outliers? I.e. value is typical on all variables but
    still the combination is unusual?
-   change across time is a special case of covariation. Explore
    temporal trends!
