---
title: "Felfedező adatelemzés"
output:
  html_notebook:
    df_print: paged
    toc: true
    toc_float:
      toc_collapsed: true
---

[link to R Notebook documentation](https://bookdown.org/yihui/rmarkdown/notebook.html)

```{r, include=FALSE}
# with this line we could set that code should not appear in the knitted document
# knitr::opts_chunk$set(echo=FALSE)
```

Let’s load a few packages!

```{r}
library(tidyverse)
library(plotly)
library(htmlwidgets)
library(GGally)
library(gridExtra)
library(corrplot)
library(patchwork)
```


# What is exploratory data analysis about?

The goal is to begin understanding the data.

A creative process — rely on your intuition too.

1. Come up with questions about the data!
2. Try to answer them using visualizations, transformations, and models!
3. Based on these, refine your questions and ask new ones!

It’s worth starting with lots of questions; later it becomes clear
which ones deserve more attention.

There are no strict rules, but two questions are usually interesting:

- what is the variance of the variables?
- what is the covariance of the variables?

## Examining the variance of a single variable

Visualizing distributions

Categorical variables: bar chart
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = cut))
```

The height of the bars shows the count of each possible *x* value  
```{r}
diamonds %>% 
  count(cut)
```


Continuous variable distributions can be viewed using a histogram
```{r}
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat), binwidth = 0.5)
```


How the displayed values are computed:

```{r}
diamonds %>% 
  count(cut_width(carat, 0.5))
```

The `binwidth` argument of `geom_histogram` determines  
how large the intervals are when splitting the x variable for plotting —  
try multiple values!

Example: only diamonds under 3 carats, with narrower bins:

```{r}
smaller <- diamonds %>% 
  filter(carat < 3)

ggplot(data = smaller, mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.1)

```


Same, faceted by cut:
```{r, fig.height=6}
ggplot(data = smaller, mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.1) + 
  facet_wrap(~ cut, nrow = 5)
```


If we want to show multiple category distributions on one plot, use lines:
```{r}
ggplot(data = smaller, mapping = aes(x = carat, colour = cut)) +
  geom_freqpoly(binwidth = 0.1, size = 2)
```


What to observe?

- What are the most common values? Why?
- What are the rarest values? Why? Did we expect this?
- Any unusual or strange patterns? What might cause them?


What’s interesting about this figure?
```{r}
ggplot(data = smaller, mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.01)
```


And this one?
```{r}
ggplot(data = faithful, mapping = aes(x = eruptions)) + 
  geom_histogram(binwidth = 0.25) + 
  xlab("Eruption duration (minutes)") +
  ylab("Frequency") + 
  ggtitle("Old Faithful geyser eruptions")
```

Clusters of similar values may indicate subgroups.

- In what ways are observations within a group similar?
- How do groups differ from each other?
- How can we characterize the clusters?
- Why might clusters be misleading?


Outliers — unusual values

Possible causes:

- data entry error  
- rare but informative event  

```{r}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5)
```

Where is the outlier? Let’s zoom in:
```{r}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5) +
  coord_cartesian(ylim = c(0, 50))
```


Let’s inspect these observations more carefully:
```{r}
(unusual <- diamonds %>% 
  filter(y < 3 | y > 20) %>% 
  select(price, x, y, z) %>%
  arrange(y))
```

Think carefully: what might have caused these anomalies?


### Practice — plotting distributions

Investigate the distribution of the `price` variable! Do you notice anything unusual?



## Missing values

What should we do with these strange diamonds?

- We can drop them — why is this not ideal?

```{r}
diamonds2 <- diamonds %>% 
  filter(between(y, 3, 20))
```


- Or we replace the strange values with missing values

```{r}
diamonds2 <- diamonds %>% 
  mutate(y = ifelse(y < 3 | y > 20, NA, y))
```


```{r}
ggplot(data = diamonds2, mapping = aes(x = x, y = y)) + 
  geom_point()
```

ggplot2 warns about NAs

But sometimes observations with missing values are exactly the interesting ones.  
For example: when were cancelled flights supposed to depart?

```{r}
nycflights13::flights %>% 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  ) %>% 
  ggplot(mapping = aes(sched_dep_time)) + 
  geom_freqpoly(mapping = aes(colour = cancelled), binwidth = 1/4)
```

Why is this figure not very informative? You'll get the chance to improve it in a bit.


## Covariance of variables


### Categorical vs continuous

```{r}
ggplot(data = diamonds, mapping = aes(x = price)) + 
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)
```

Because the frequencies differ greatly, we cannot clearly see  
whether distributions differ across cuts.

```{r}
ggplot(diamonds) + 
  geom_bar(mapping = aes(x = cut))
```


Instead of counts, show density:
```{r}
ggplot(data = diamonds, mapping = aes(x = price, y = ..density..)) + 
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)
```


Or use a boxplot:
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, y = price)) +
  geom_boxplot()
```

Are lower-quality diamonds more expensive?

Cut is ordinal, but not all categorical variables are.  
For example, fuel efficiency across car types:

```{r}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) +
  geom_boxplot()
```

The plot becomes clearer if we sort classes by median highway mileage:

```{r}
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy))
```

If variable names are long, flip the plot:
```{r}
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy)) +
  coord_flip() + 
  xlab("Class") + 
  ylab("Highway miles per gallon") + 
  ggtitle("Classes of cars ranked by highway mileage")
```

#### Practice — covariance visualization 1

How could we improve the following figure based on what we've learned?

```{r}
nycflights13::flights %>% 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  ) %>% 
  ggplot(mapping = aes(sched_dep_time)) + 
  geom_freqpoly(mapping = aes(colour = cancelled), binwidth = 1/4)
```


### Categorical vs categorical

```{r}
ggplot(data = diamonds) +
  geom_count(mapping = aes(x = cut, y = color))
```

Frequency table:
```{r}
diamonds %>% 
  count(color, cut) 
```

Show as a heatmap:
```{r}
diamonds %>% 
  count(color, cut) %>%  
  ggplot(mapping = aes(x = color, y = cut)) +
  geom_tile(mapping = aes(fill = n))
```

#### Practice — covariance visualization 2

Using dplyr and geom_tile(), examine average delay by destination and month.  
Why might this figure be difficult to interpret?  
What could we do with it?




### Continuous vs continuous

```{r}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price))
```

Try transparency:
```{r}
ggplot(data = diamonds) + 
  geom_point(mapping = aes(x = carat, y = price), alpha = 1 / 100)
```

Transparency doesn’t always help with large datasets.  
In that case, bin the data:

```{r}
ggplot(data = smaller) +
  geom_bin2d(mapping = aes(x = carat, y = price), bins = 30)
```

Or slice only one variable:
```{r}
ggplot(data = smaller, mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes(group = cut_width(carat, 0.1)))
```

To show sample sizes in each box:
```{r}
ggplot(data = smaller, mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes(group = cut_width(carat, 0.1)), varwidth = TRUE)
```


Or slice into equal-sized bins:
```{r}
ggplot(data = smaller, mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes(group = cut_number(carat, 20)))
```

#### Practice — covariance visualization 3

Using what we’ve learned, visualize the joint distribution of cut, carat, and price!



# Complex exploratory analysis quickly

## GGally

Show distributions and relationships in one large figure:
```{r, fig.width=10, fig.height=10, message=FALSE}
ggpairs(iris, progress=FALSE)
```

Base R version:
```{r, fig.height=7,  fig.width=7}
pairs(iris, pch=16)
```

Additional scatterplot matrix options:  
http://www.sthda.com/english/wiki/scatter-plot-matrices-r-base-graphs#r-base-scatter-plot-matrices-pairs

`ggcoef()` is useful for quick inspection of model coefficients.

Example:

Which variables relate to fuel consumption?
```{r}
mtcars %>% 
  select(mpg, disp, hp, wt) %>% 
  pairs()
```

Fit a multiple regression model:
```{r}
model <- 
  mtcars %>% 
  select(mpg, disp, hp, wt) %>% 
  # standardize variables to get standardized coefficients
  mutate_all(scale) %>% 
  lm(mpg ~  ., data = .)

summary(model)
```

```{r}
confint(model)
```

Plot coefficients:
```{r}
ggcoef(model)
```


## corrplot

Correlation matrices:

```{r}
(corr_mat_cars <- 
   mtcars %>% 
   select(mpg, disp, hp, wt) %>% 
   cor())
```


```{r}
corr_mat_cars %>% 
  corrplot()
```

```{r}
p_mat <- 
  mtcars %>% 
  select(mpg, disp, hp, wt) %>% 
  cor.mtest() %>% .[["p"]]

corr_mat_cars %>% 
  corrplot(p.mat       = p_mat, 
           method      = "square", 
           addCoef.col = "lightgrey", 
           mar         = c(0.2,1,1,0.5), 
           insig       = "blank")
```

corrplot is especially useful with many variables.

Example with Big Five personality test (25 items):
```{r}
psych::bfi %>% 
  select(-(gender:age)) %>% 
  cor(use = "p") %>% 
  corrplot(method      = "square", 
           mar         = c(0.2,1,1,0.5))
```


# Multi-panel figures

## gridArrange package

```{r, message=FALSE, fig.width=15}
p1 <- ggplot(diamonds) + 
  geom_freqpoly(aes(price), col="blue",binwidth = 1000)

p2 <- ggplot(diamonds) + 
  geom_point(aes(carat, price), alpha = 0.01, col="blue")

p3 <- ggplot(diamonds) + 
  geom_histogram(aes(price), fill="blue", binwidth = 10)

grid.arrange(p1, p2, p3, nrow = 1)
```

Fine-tune layout via `layout_matrix`:

```{r, fig.width=15}
grid.arrange(
  p1, p2, p3, 
  layout_matrix = 
    rbind(
      c(1, 2, 2),
      c(3, 2, 2))
)
```

https://cran.r-project.org/web/packages/egg/vignettes/Ecosystem.html

## patchwork package

```{r}
p1 + p2 + p3 + plot_layout(ncol = 2)
```


# Interactive graphics with plotly

https://plotly-r.com/

Volcano example:
```{r}
plot_ly(z = ~volcano)
```


```{r}
plot_ly(z = ~volcano) %>% add_surface()
```


Texas real estate data:
```{r}
txhousing
```


Two approaches to plotly graphics:

1) `ggplotly()` — convert a ggplot  
2) `plot_ly()` — specify data directly

## ggplotly(): converting ggplot

```{r}
(p <- ggplot(txhousing, aes(date, median)) +
  geom_line(aes(group = city), alpha = 0.2))
```

Tooltip shows all aesthetics:
```{r}
ggplotly(p)
```

We can customize tooltip:
```{r}
ggplotly(p, tooltip = "city")
```


## plot_ly(): direct data specification

Use dplyr/tidyr for data manipulation.  
plotly understands grouped data (creates elements per group).

```{r}
tx <- group_by(txhousing, city)
```

Create plotly object:
```{r}
p <- plot_ly(tx, x = ~date, y = ~median)
```

Inspect data:
```{r}
plotly_data(p)
```

Add lines:
```{r}
p %>% 
  add_lines(alpha = 0.2, name = "Texan Cities", hoverinfo = "none") %>% 
  add_lines(name = "Houston", data = filter(txhousing, city == "Houston"))
```

`add_*()` functions:

- inherit `plot_ly()` axis & aesthetic settings  
- inherit the plotly object's data (can be overridden via `data=`)

Think of them like ggplot2 layers.

